---
title: "SOTB-Analysis-JT"
output: html_document
date: "2026-01-26"
---

This is the final analysis document for the 2025 SOTB report - Appendix D - Monitoring Results.

Each Indicator will have its own section for metrics, algorithms (not included in final report) and figures.

Metrics downloaded from (1) EMPA data portal, (2) CRAM data portal, (3) from CW

List of indicators (each has its own algorithm described in Appendix A):

CONDITION INDICATORS
- Trophic Food Web Support
- Nursery Habitat Provisioning for Fish
- Forage and Breeding Function for Birds
- Habitat Structure and Complexity
- Vegetation Community
- Sea-level rise Amelioration

STRESSOR INDICATORS
- Eutrophication
-Anthropogenic Disturbance and Land Use
-Altered Hydrology
-Non-native Species
-Contaminant & Excessive Sediment Loading


We will include the following data:

Year - 2024
  If data is available from 2023, we will include it the clean datasets for later filtering
Season - Spring and Fall
Sites - 
  Malibu Lagoon
  Topanga Canyon
  Zuma Lagoon
  Arroyo Sequit
  Big Sycamore 


#Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Open Libs
```{r Packages}

library(tidyr)
library(dplyr)
library(tidyverse)
library(readr)
library(stringr)
library(clipr)
library(lubridate)
library(reshape2)
library(vegan)
library(ggplot2)
library(ggrepel)
library(here)
library(ggpubr)
library(openxlsx)

library(viridis)
library(RColorBrewer)
library(pals)

```
# Step 1: Load all datasets and clean data

```{r source loading script, cache = T}
source(here("SM-Bay-scripts", "01_load_data.R"))
source(here("SM-Bay-scripts", "02_clean_data.R"))
```


# Step 3: Monitoring data figures

 Fish Metrics
 
```{r fish monitoring data - species richness, heatmap, nmds, echo = F}

#site housekeeping
## add order of sites (north to south)
site_order <- c("SC-BSC", "SC-ASE", "SC-ZUM","SC-TOP", "SC-MAL")

## define colors for color category
colors <- c("Spring" = "#DDCC77", "Fall" = "#88CCEE", "Both" = "#117733",
            "None" = "white")
#heatmap

fish_heatmap <- fish_filtered_data %>% 
  ggplot(aes(commonname, siteid, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Fish Species", y = "SiteID", title = "2024 Fish Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
fish_heatmap

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/fish_heatmap.jpg",
           fish_heatmap,
       width = 12,
       height = 8,
           dpi = 300)


#nMDS
str(PA_filter)

Fish_PA_data_wide <- PA_filter %>% 
  #select variables for comparison
  select(siteid, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season,
         commonname, presence_absence) %>%
  #get rid of sites with Not recorded fish, therefore getting rid of the sites with no fish
  filter(commonname != "Not recorded") %>% 
  #Switch to wide format
  pivot_wider(names_from = commonname, values_from = presence_absence, values_fill = 0 ) %>%
  unite("SiteID_full", siteid, stationno, netreplicate, date,
        sep = "-",
        remove = FALSE ) #%>% 
  #filter(SiteID_full != "SC-VEN-3-2-2023-11-03")

Fish_meta_df <- Fish_PA_data_wide %>% 
  select(siteid, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season) 

Fish_matrix <- Fish_PA_data_wide %>% 
  select(-c(siteid, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season)) %>% 
  column_to_rownames("SiteID_full")

#Permanvoa

Fish_perm <- adonis2(Fish_matrix ~ season,
                data = Fish_meta_df)
Fish_perm


#nmds

nmds.Fish <- metaMDS(Fish_matrix, distance = "bray", autotransform = TRUE)
#autotransform = TRUE for transformation, otherwise keep FALSE
nmds.Fish #stress = 0.112 1/27/26

plot(nmds.Fish)

#look for outliers
orditorp(nmds.Fish, "sites") 
#identify the outliers in order to drop from analysis = SC-BAT-1-3

# Stress Plot
stressplot(nmds.Fish, main = "Shepard plot")

# Extract scores

# Extract site scores (samples)
site.scores <- as.data.frame(scores(nmds.Fish, display = "sites"))
site.scores$SampleID <- rownames(site.scores)  # optional for labeling

# Extract species scores (species vectors)
species.scores <- as.data.frame(scores(nmds.Fish, display = "species"))
species.scores$Species <- rownames(species.scores)

# Add metadata to sites
site.scores <- cbind(site.scores, Fish_meta_df)

####extract signficant species
#this can be repeated for enironmental vectors
#replicae "vectors" with "factors"

# Fit species to the NMDS ordination
fish_envfit <- envfit(nmds.Fish, Fish_matrix, permutations = 999)

# Extract p-values
species_pvals <- as.data.frame(fish_envfit$vectors$pvals)
species_scores <- as.data.frame(scores(fish_envfit, display = "vectors"))
species_r <- as.data.frame(fish_envfit$vectors$r)

# Combine
species_scores$Species <- rownames(species_scores)
species_scores$pval <- species_pvals[, 1]
species_scores$r <- species_r[, 1]

# Keep only p < 0.01
species_sig <- species_scores %>%
  filter(pval < 0.01)

#Rescale NMDS vectors by r
species_sig <- species_sig %>%
  mutate(NMDS1_scaled = NMDS1 * r * 3,
         NMDS2_scaled = NMDS2 * r * 3)


# PLot contains all sites - not a big deal

nMDS.plot.fish <- site.scores %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot() +
  # Sample points
  geom_point(data = site.scores, aes(x = NMDS1, y = NMDS2, 
                                     color = season,
                                     shape = siteid), 
             size = 3) +
  
  #Confidence ellipses
  stat_ellipse(data = site.scores,
               aes(x = NMDS1, y = NMDS2, color = season),
               type = "t", level = 0.95, linetype = "dashed", linewidth = 1) +
  #type = "t" uses a multivariate t-distribution (better for small sample sizes).
  #level = 0.95 is for 95% confidence ellipses.
  #You can switch to type = "norm" if assumptions for t-distribution don’t hold.

  # Arrows scaled by r
  geom_segment(data = species_sig %>%
                 arrange(desc(r)) %>%   # Sort descending by r
                 slice_head(n = 5),      # Take top 5 rows,
               aes(x = 0, y = 0, xend = NMDS1_scaled, yend = NMDS2_scaled),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +

  #Species labels at arrow tips
  geom_text_repel(
    data = species_sig %>%
                 arrange(desc(r)) %>%   # Sort descending by r
                 slice_head(n = 5),
    aes(x = NMDS1_scaled, y = NMDS2_scaled, label = Species),
    size = 4, fontface = "bold", color = "black",
    nudge_y = 0.05, # small nudge to prevent overlap with arrow tip
    segment.size = 0.5,
    segment.color = "grey50"
  ) +
  
  #aesthetics 
  theme_bw() + 
  labs(title = "NMDS with Significant Species Vectors (Scaled by r)",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)) +
  coord_fixed() +  ## need aspect ratio of 1
  
  #add stress label
  annotate(geom = "label", x = 5, y = -1, size = 3,
           label = paste("Stress: ", round(nmds.Fish$stress, digits = 3))) +
  
  #add specifications for colors and shapes
  scale_shape_manual(name = "Site",
                     values=c(15, 16, 17, 18, 3, 4, 8)) +
  scale_color_manual(name = "Season",
                       values = c("#DDCC77", "#88CCEE"),
                     breaks = c("Spring", "Fall"))

nMDS.plot.fish


ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Fish_nMDS_species.jpg", 
           nMDS.plot.fish,
       width = 10,
       height = 6,
           dpi = 300)

# First Calculate the total number of nets for each site
# Summarize the number of unique combinations of netreplicate and stationno for each site
netreplicate_data <- PA_filter %>% #we will use PA_filter because this includes all nets
  group_by(siteid, season) %>%
  summarize(total_replicates = n_distinct(paste(netreplicate, stationno)), .groups = 'drop')

#Calculate SR by site
Fish_PA_data_wide_SR_site <- Fish_PA_data_wide %>% 
  group_by(siteid, estuaryname, region,
         estuaryclass, mpastatus, estuarytype, year, season) %>% 
  summarise(across('California halibut':'Western mosquitofish', sum)) %>% 
  ungroup() %>% 
  #change numeric values to 1 so that it's PA
  mutate(across(c('California halibut':'Western mosquitofish'), ~ if_else(. > 0, 1, 0))) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'California halibut':'Western mosquitofish') %>% 
      rowSums()) %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  #add in 0 for species richness
  mutate(species_richness = ifelse(is.na(species_richness), 0, species_richness)) %>% 
  left_join(., netreplicate_data) %>% 
  #Arroyo Sequit Fall was not fished, so need to change the 0 back to NA
  mutate(species_richness = case_when(
    siteid == "SC-ASE" & season == "Fall" ~ NA,
    TRUE ~ species_richness
  ))


SR_site_plot <- Fish_PA_data_wide_SR_site %>% 
  mutate(siteid = factor(siteid, levels = site_order)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid, y=species_richness, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Fish species richness") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) + 
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  geom_text(aes(label = total_replicates), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5)   # Show total replicates as label

SR_site_plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Fish_SpeciesRichness_bysite.jpg", 
           SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

```
 
 Fish Length Metrics

```{r fish length monitoring data, include = F, echo = F}
str(Length_Data_clean)

#calculate the range of fish sizes

Fish_size_range <- Length_Data_clean %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  #select only fish
  filter(fish_or_invert == "fish") %>% 
  filter(year == "2024") %>% 
  group_by(siteid, season,
           estuaryname, estuarytype, year) %>% 
  summarise(fish_range = max(length_mm) - min(length_mm))

#lets plot a full boxplot of all the lengths 

Fish_size_boxplot <- Length_Data_clean %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x = siteid, y = length_mm, fill = season)) +
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), width = 0.6) +
  geom_jitter(aes(color = season, group = season),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  scale_color_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  labs(x = "Site", y = "Fish length (mm)") +
  scale_x_discrete(limits = (levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom"
  ) +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) 

Fish_size_boxplot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Fish_SizeRange_bysite.jpg", 
           Fish_size_boxplot,
       width = 10,
       height = 8,
           dpi = 150)


```
#Invertebrate Metrics

```{r invert monitoring data, include = F, echo = F}
str(PA_all)

# invert species heatmap

n_species_invert <- length(unique(PA_filter_order_invert$scientificname))

heatmap_large_invert <- PA_filter_order_invert %>% 
  ggplot(aes(scientificname, siteid, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Macro-invertebrate species", y = "SiteID", title = "2024 Macro-Invertebrate Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order_invert$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species_invert - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap_large_invert

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Macro_Invert_PA_heatmap.jpg", 
           heatmap_large_invert,
       width = 12,
       height = 8,
           dpi = 150)

```

Temperature and Dissolved Oxygen Metrics

```{r temperature and DO, echo = F}
do_plot <- ggplot(temp_do) +
  geom_bar(aes(x = siteid, y = num_events_DO), stat = "identity", fill = "darkred") +
  ylab("number of annual DO events < 3 mg/L") +
  xlab("estuary") +
  theme_bw() +
  facet_wrap(~season) +
  theme(axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size =24),
        axis.text.x = element_text(size = 18, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, angle = 45, hjust = 1),
        strip.text = element_text(size = 20)) 
  
 
do_plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/DO_plot.jpg",
       do_plot, 
       height = 12,
       width = 12,
       dpi = 300)

temp_plot <- ggplot(temp_do) +
  geom_bar(aes(x = siteid, y = perc_temp_abv_25), stat = "identity", fill = "red") +
  ylab("average percent time above 25C") +
  xlab("estuary") +
  theme_bw() +
  ylim(c(0,100)) +
  facet_wrap(~season) +
   theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size =18),
        axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 16, angle = 45, hjust = 1),
        strip.text = element_text(size = 18)) 
  
temp_plot

combined_stressor <- cowplot::plot_grid(do_plot, temp_plot, labels = c("A", "B"), label_size = 16, ncol = 1)
combined_stressor

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/temp_do_plot.jpg",
       combined_stressor, 
       height = 12,
       width = 12,
       dpi = 300)
```
#Bird Metrics

```{r birds, echo = F}
str(bird_count_df)

PA_birds <- bird_count_df %>%
  mutate(presence_absence = ifelse(abundance != 0, 1, 0)) %>% 
  #malibu was sampled twice in spring 2024 - select commmon month
  filter(month != 5)


## find order of PA from most to least for ordering of heat map
SpeciesOrder_bird <- PA_birds %>%
  filter(presence_absence == 1) %>% 
  select(commonname, siteid) %>% 
  distinct() %>% 
  group_by(commonname) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

## get the ordered list of species by their count
SpeciesList_bird <- SpeciesOrder_bird %>% pull(commonname)

## summarize df and create color category for PA heat map. will add new color if present in both years 
PA_summarized_bird <- PA_birds %>%
  group_by(siteid, commonname) %>%
  summarise(color_category = case_when(
    sum(presence_absence == 1 & season == "Fall") > 0 & 
      sum(presence_absence == 1 & season == "Spring") > 0 ~ "Both",
    sum(presence_absence == 1 & season == "Spring") > 0 ~ "Spring",
    sum(presence_absence == 1 & season == "Fall") > 0 ~ "Fall",
    TRUE ~ "None"
  )) %>%
  ungroup() #

setdiff(unique(PA_summarized$commonname), SpeciesList) 
#more species are showing up then in the list because they are 0's
#we should filter them out

## order based on site order
PA_filter_order_bird <-PA_summarized_bird %>% 
  mutate(siteid =  factor(siteid, levels = site_order)) %>%
  arrange(siteid)  %>% 
  #order based on species count
  filter(commonname %in% SpeciesList_bird) %>% 
  mutate(commonname =  factor(commonname, levels = SpeciesList_bird)) %>%
  arrange(commonname)  


##GGPLOT

n_species_bird <- length(unique(PA_filter_order_bird$commonname))


heatmap_birds <- PA_filter_order_bird %>% 
  ggplot(aes(commonname, siteid, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Bird Species", y = "SiteID", title = "2024 Bird Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order_bird$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species_bird - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap_birds


ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Bird_PA_heatmap.jpg", 
           heatmap_birds,
       width = 16,
       height = 6,
           dpi = 150)



```
#Vegetation Metrics

```{r vegetation, echo = F}
str(PA_veg)
str(PA_filter_order_veg)

##GGPLOT

n_species_veg <- length(unique(PA_filter_order$scientificname))

heatmap_veg <- PA_filter_order %>% 
  ggplot(aes(scientificname, siteid_star, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Plant Species", y = "SiteID", title = "2024 Plant Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species_veg - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap_veg

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Veg_PA_heatmap.jpg", 
           heatmap_veg,
       width = 20,
       height = 6,
           dpi = 250)
```

# CRAM Metrics

```{r CRAM, echo = F}

#using only 2022 data for SOTB report  - downloaded 11/20/25

str(CRAM_data_clean)

CRAM_index <- CRAM_data_clean %>% 
  group_by(siteid) %>% 
  summarise(index_avg = mean(indexscore),
            buffer_avg = mean(buffer_landscape_cont),
            hydrology_avg = mean(hydrology),
            physical_structure_avg = mean(physical_structure),
            biotic_structure_avg = mean(biotic_structure)) %>% 
  pivot_longer(cols = c(index_avg:biotic_structure_avg), names_to = "index")

index_labels <- c(
  buffer_avg = "Buffer",
  hydrology_avg = "Hydrology",
  physical_structure_avg = "Physcial Structure",
  biotic_structure_avg = "Biotic Structure",
  index_avg = "Index")


CRAM_plot_subindex <- CRAM_index %>% 
  filter(index != "index_avg") %>% 
  ungroup() %>% 
  #complete(siteid, year_label, index, fill = list(value = NA, vyear = NA)) %>% 
  mutate(siteid = factor(siteid, levels = site_order)) %>% 
 # mutate(year_label = factor(year_label, levels = c("year 1", "year 2"))) %>% 
  ggplot(aes(x=siteid, y=value)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Score") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~index,
             labeller=labeller(index = index_labels)) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Survey",
                    values = c("#DDCC77", "#88CCEE"),
                    labels = c("Prior", "Recent")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) #+
  # geom_text(aes(label = vyear),
  #           position = position_dodge(width = 0.9),
  #           vjust = -0.5, size = 3.5,
  #           na.rm = TRUE)   # Show total replicates as label

CRAM_plot_subindex

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/CRAM_subindices_2022.jpg", 
           CRAM_plot_subindex,
       width = 10,
       height = 8,
           dpi = 150)


# overall index plot - 2022 data only, folks

CRAM_plot <- CRAM_index %>% 
  filter(index == "index_avg") %>% 
  ungroup() %>% 
  #complete(siteid, index, fill = list(value = NA, vyear = NA)) %>% 
  mutate(siteid = factor(siteid, levels = site_order)) %>% 
  #mutate(year_label = factor(year_label, levels = c("year 1", "year 2"))) %>% 
  ggplot(aes(x=siteid, y=value)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Score") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~index,
             labeller=labeller(index = index_labels)) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
 # scale_fill_manual(name = "Survey",
      #              values = c("#DDCC77", "#88CCEE"),
      #              labels = c("Prior", "Recent")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) #+
  # geom_text(aes(label = vyear),
  #           position = position_dodge(width = 0.9),
  #           vjust = -0.5, size = 3.5,
  #           na.rm = TRUE)   # Show total replicates as label

CRAM_plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/CRAM_index.jpg", 
           CRAM_plot,
       width = 10,
       height = 8,
           dpi = 150)


```

# Macroalgae metrics
```{r macroalgae, echo = F}

Floating_plot <- Macro_Floating_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, estuarytype,
           year, region, season) %>% 
  summarise(Macro_avg = mean(estimatedcover, na.rm = TRUE),
            Macro_SD = sd(estimatedcover, na.rm = TRUE)) %>% 
  #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = rev(site_order))) %>% 
  #mutate(year = factor(year, levels = c("2023", "2024"))) %>% 
  ungroup() %>%  # Ungroup the data before completing
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid, year, season, fill = list(Macro_avg = NA, Macro_SD = NA))  %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid, y=Macro_avg, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(0.9)) +
  #prevent the error bars from going below 0
  geom_errorbar(aes(ymin= pmax(Macro_avg-Macro_SD, 0), ymax=Macro_avg+Macro_SD), 
                width=0.2, position= position_dodge(0.9)) +
  ylab("Avg. % Floating Macroalgae Cover") + 
  xlab("Site") +
  theme_bw() +
  #facet_wrap(~year) %>% 
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  coord_flip() 


Floating_plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Macro_floating.jpg", 
           Floating_plot,
       width = 10,
       height = 8,
           dpi = 150)


```


# Step 4: Indicator Figures and Algorithms

# Trophic Food Web Support (modified from ideal)

trophic indicator = (fish abundance + invert abundance)/2

```{r trophic food web support indicator, echo = F}
str(sotb_invert_data_clean)

# inverts
invert_abun <- sotb_invert_data_clean %>%
  filter(!is.na(scientificname), scientificname != "-88") %>%
  group_by(siteid, year, season) %>%
  summarize(invert_abundance = sum(abundance), .groups = "drop") 

invert_richness <- sotb_invert_data_clean %>%
  filter(!is.na(scientificname), scientificname != "-88") %>%
  group_by(siteid, year, season) %>%
  summarize(invert_richness = n_distinct(scientificname), .groups = "drop")

# feesh
str(Seine_df)

fish_abun <- Seine_df %>%
  filter(!is.na(abundance), abundance != "-88") %>%
  group_by(siteid, year, season) %>%
  summarize(fish_abundance = sum(abundance), .groups = "drop")

fish_richness <- Seine_df %>%
  filter(!is.na(scientificname), scientificname != "-88") %>%
  group_by(siteid, year, season) %>%
  summarize(fish_rich = n_distinct(scientificname), .groups = "drop")

#Trophic Support Metric

trophic_combined <- invert_abun %>%
  left_join(fish_abun, by = c("siteid", "year", "season")) %>%
  mutate(trophic_support_indicator = (fish_abundance + invert_abundance)/2) %>%
  mutate(trophic_support_indicator = as.integer(trophic_support_indicator))
  

Trophic_Support_metric <- trophic_combined %>%
  pivot_longer(cols = c(invert_abundance, fish_abundance, trophic_support_indicator), 
               names_to = "metric", values_to = "value") 
  
# tile plot with all components + algorithm
Trophic_Support_metric$metric <- factor(recode(Trophic_Support_metric$metric, 
                             fish_abundance = "Fish Abundance",
                             invert_abundance = "Invert Abundance",
                             trophic_support_indicator = "Trophic Support Indicator"), 
                             levels = c("Fish Abundance", "Invert Abundance", "Trophic Support Indicator"))

trophic_colors <- c(`Fish Abundance` = "lightblue", `Invert Abundance` = "lightblue", `Trophic Support Indicator` = "cornflowerblue")

Trophic_Support_Indicator_Plot <- Trophic_Support_metric %>%
  filter(year %in% 2024) %>%
  ggplot( aes(metric, siteid, fill = metric)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), size = 8) +
  facet_wrap(~season) +
  scale_fill_manual(values = trophic_colors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 18)) +
  theme(legend.position = "none")

Trophic_Support_Indicator_Plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Trophic_Support_indicator.jpg", 
           Trophic_Support_Indicator_Plot,
       width = 12,
       height = 12,
           dpi = 200)

# Tile Plot for Trophic Food Web Support

```

# Nursery Habitat Provisioning for Fish


# Bird Habitat Support (modified from ideal)
bird indicator = bird abundance + bird richness

```{r bird indicator, include = F}
str(bird_count_df)
str(PA_birds)
#Metrics included in this indicator are abundance and richness

#create df of site, year, season, metric  and value 
bird_abun <- bird_count_df %>%
  group_by(siteid, year, season) %>%
  summarize(bird_abundance = sum(abundance), .groups = "drop")

bird_richness <- PA_birds %>%
  group_by(siteid, year, season) %>%
  summarize(bird_richness = sum(presence_absence), .groups = "drop") 

# combined bird dataframe + algorithm for bird indicator
bird_combined <- bird_abun %>%
  left_join(., bird_richness, by = c("siteid", "year", "season")) %>%
  group_by(siteid, year, season) %>%
  mutate(bird_indicator = bird_abundance + bird_richness)

Bird_metric <- bird_combined %>%
  pivot_longer(cols = c(bird_abundance, bird_richness, bird_indicator), 
               names_to = "metric", values_to = "value")

  
# tile plot with all components + algorithm
Bird_metric$metric <- factor(recode(Bird_metric$metric, 
                             bird_abundance = "Bird Abundance",
                             bird_richness = "Bird Richness",
                             bird_indicator = "Bird Indicator"), 
                             levels = c("Bird Abundance", "Bird Richness", "Bird Indicator"))

bird_colors <- c(`Bird Abundance` = "lightblue", `Bird Richness` = "lightblue", `Bird Indicator` = "cornflowerblue")

Bird_Indicator_Plot <- ggplot(Bird_metric, aes(metric, siteid, fill = metric)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), size = 8) +
  facet_wrap(~season) +
  scale_fill_manual(values = bird_colors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 18)) +
  theme(legend.position = "none")

Bird_Indicator_Plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figures_share/Bird_indicator.jpg", 
           Bird_Indicator_Plot,
       width = 12,
       height = 12,
           dpi = 200)

```

# Habitat Structure and Complexity
# Vegetation Community
# Eutrophication
# Anthropogenc Disturbance and Land Use
# Altered Hydrology
# Non-native Species

# BPJ scores

Figures summarizing Best Professional Judgment exercise by Expert Panel members

```{r BPJ score summary}


bpj.scores <- here(read.csv("data-inventory", "BPJ_scores.csv"))
head(bpj.scores)

# Pivot data
bpj.scores.long <- bpj.scores %>%
  pivot_longer(cols = c(Avg.Rnd.1, Avg.Rnd.2),
               names_to = "Round",
               values_to = "Average") %>%
  mutate(stdev = ifelse(Round == "Avg.Rnd.1", stdev_R1, stdev.2_R2))

round.colors = c("Avg.Rnd.1" = "darkblue", "Avg.Rnd.2" = "#00cCCC")
class.colors = c("seasonally dry, small lagoons" = "#330000", "sandy beach, usually wet lagoons" = "#FFCCCC" , "larger lagoons, usually wet" = "#990000")

# Plot 1: plot just stdev in both rounds

scoring_summary1 <- ggplot(bpj.scores.long, aes(x = Estuary, y = Average, color = Round))  + 
  geom_tile(aes(y = 0.5, fill = Class),
            height = 0.3, color = NA) +
  geom_point(size =  3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = Average - stdev,
                    ymax = Average + stdev),
                width = 0.1, position = position_dodge(width = 0.4))  +
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.05)), breaks = 1:5) +
  scale_color_manual(values = round.colors, labels = c("Avg.Rnd.1" = "Round 1", "Avg.Rnd.2" = "Round 2"), name = "Panel Review") +
  scale_fill_manual(values = class.colors, name = "Estuary class") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 14),
        legend.position = "right") + 
  labs(y = "Score 1 -5", x = "Estuary") 
  
scoring_summary1

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/SOTB_ppt/scoring_summary1.jpg", 
           scoring_summary1,
       width = 8,
       height = 8,
           dpi = 300)


# Plot 2: Plot side-by side barplot with consensus 
#repivot data to include consensus
bpj.scores.consensus <- bpj.scores %>%
  pivot_longer(cols = c(Avg.Rnd.1, Avg.Rnd.2, Consensus),
               names_to = "Round",
               values_to = "Average") %>%
  mutate(stdev = case_when(
    Round == "Avg.Rnd.1" ~ stdev_R1,
    Round == "Avg.Rnd.2" ~ stdev.2_R2,
    Round == "Consensus" ~ NA_real_
  ))

round.colors = c("Avg.Rnd.1" = "darkblue", "Avg.Rnd.2" = "#00cCCC", "Consensus" = "#339933")
class.colors = c("seasonally dry, small lagoons" = "black", "sandy beach, usually wet lagoons" = "grey90" , "larger lagoons, usually wet" = "grey50")


consensus_summary1 <-  ggplot(bpj.scores.consensus, aes(x = Estuary, y = Average, color = Round))  + 
  geom_tile(aes(y = 0.5, fill = Class),
            height = 0.3, color = NA) +
  geom_point(size =  3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = Average - stdev,
                    ymax = Average + stdev),
                width = 0.1, position = position_dodge(width = 0.4))  +
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.05)), breaks = 1:5) +
  scale_color_manual(values = round.colors, labels = c("Avg.Rnd.1" = "Round 1", "Avg.Rnd.2" = "Round 2", "Consensus" = "Consensus"), name = "Panel Review") +
  scale_fill_manual(values = class.colors, name = "Estuary class") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 14),
        legend.position = "right") + 
  labs(y = "Score 1 -5", x = "Estuary") 

  
  consensus_summary1
  
  ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/SOTB_ppt/scoring_summary2.jpg", 
           consensus_summary1,
       width = 8,
       height = 8,
           dpi = 300)

# Slide 23 replacement

```
