---
title: "SOTB-Analysis-JW"
author: "Jan Walker"
date: "2025-06-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

We will use the Bight 2023 analyses as a template to begin compiling the data for the SOTB Report.

Each Indicator will have its own section for metrics and figures.

A final table will pull the data together.

We will include the following data:
Year - 2024
  If data is available from 2023, we will include it the clean datasets for later filterin
Season - Spring and Fall
Sites - 
  Goleta Slough
  Ventura River
  Malibu Lagoon
  Newport Bay
  Batiquitos Lagoon

  Topanga Canyon
  Zuma Lagoon
  Arroyo Sequit
  Big Sycamore 


#Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Packages}

library(tidyr)
library(dplyr)
library(tidyverse)
library(readr)
library(stringr)
library(clipr)
library(lubridate)
library(reshape2)
library(vegan)
library(ggplot2)
library(ggrepel)

library(ggpubr)
library(openxlsx)

library(viridis)
library(RColorBrewer)
library(pals)

```

```{r Sites}

#create a dataframe that we can make sure has all the right information for all sites

#add Arroyo Sequit Fall
new_row <- data.frame(
  siteid = "SC-ASE",
  estuaryname = "Arroyo Sequit",
  estuarytype = "Temporarily Closed",
  year = 2024,
  season = "Fall")

#call in Seine_df from below**

Site_df <- Seine_df %>% 
  select(siteid, estuaryname, estuarytype, year, season) %>% 
  unique() %>% 
  filter(year == 2024) %>% 
  #add a row for arroyo sequit
  rbind(., new_row) %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, season)

```


#Fish

```{r Data}

SOP_9_fish_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_fish_abundance_data")

Seine_Data_clean <- SOP_9_fish_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, method,
         stationno, netreplicate, scientificname, commonname,
         abundance, region:estuarytype, status, surveytype) %>% 
  ## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6, 7) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month)

## add lookup list for fish or invert 
fish_lookup <- read.csv("/Users/janw/Desktop/projects/Bight-Estuary-2023/Files/lu_fishmacrospecies.csv")

fish_lookup <- fish_lookup %>% 
  select(scientificname, fish_or_invert)

Seine_df <- Seine_Data_clean %>% 
  left_join(.,fish_lookup) %>% 
  #add 'zero' fish or invert type to note when species = Not recorded
  mutate(fish_or_invert = 
           ifelse(fish_or_invert == '', "zero", fish_or_invert)) %>% 
  #select only Bight data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))


##Call in length data

SOP_9_fish_length_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_fish_length_data")

Length_Data_clean <- SOP_9_fish_length_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date,
         stationno, netreplicate, scientificname, commonname, length_mm, replicate,
         region:estuarytype, status, surveytype) %>% 
  ## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6, 7) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  #remove lengths that are -88
  filter(length_mm != -88) %>% 
  select(-month) %>% 
  left_join(.,fish_lookup) %>% 
  #add 'zero' fish or invert type to note when species = Not recorded
  mutate(fish_or_invert = 
           ifelse(fish_or_invert == '', "zero", fish_or_invert)) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))



```

```{r Data inventory}

#we want to know which years and seasons we have data for

df_inventory <- Seine_df %>%
  distinct(siteid, year, season) %>%
  arrange(siteid, year, season)

df_inventory_wide <- df_inventory %>%
  mutate(present = TRUE) %>%
  distinct(siteid, year, season, present) %>%
  pivot_wider(names_from = c(year, season), values_from = present, values_fill = FALSE)

#a lot of missing data in 2023, so will probably ignore 2023

```


## Community

```{r heatmap PA}

#create a dataframe that is just Presence/Absence data
PA_Fish <- Seine_df %>%
  mutate(presence_absence = ifelse(abundance != 0, 1, 0))

## filter fish/invert and clean
PA_filter <- PA_Fish %>% 
  #add star so can flag in graphs where sites are open
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  #select only fish
  #by keeping zero - keeps in the sites with no fish
  filter(fish_or_invert == "fish" |
           fish_or_invert == "zero") %>% 
  #filter(commonname != "Not recorded") %>%
  #remove unknown fish
  filter(!grepl("unknown", commonname, ignore.case = TRUE)) %>% 
  #remove juvenile fish
  filter(!grepl("juvenile", commonname, ignore.case = TRUE)) %>% 
  filter(year == "2024")

## find order of PA from most to least for ordering of heat map
SpeciesOrder <- PA_filter %>%
  #filter(presence_absence == 1) %>% 
  select(commonname, siteid) %>% 
  distinct() %>% 
  group_by(commonname) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

## get the ordered list of species by their count
SpeciesList <- SpeciesOrder %>% pull(commonname)

## add order of sites (north to south)
site_order <- c("SC-BSC", "SC-ASE", "SC-ZUM","SC-TOP", "SC-MAL")

site_order_star <- c("SC-GOL*", "SC-VEN*", "SC-BSC*", "SC-ASE*", "SC-ZUM*",
                "SC-MAL*", "SC-TOP*", "SC-NEW", "SC-BAT")

## summarize df and create color category for PA heat map. will add new color if present in both years 
PA_summarized <- PA_filter %>%
  group_by(siteid, siteid_star, commonname) %>%
  summarise(color_category = case_when(
    sum(presence_absence == 1 & season == "Fall") > 0 & 
      sum(presence_absence == 1 & season == "Spring") > 0 ~ "Both",
    sum(presence_absence == 1 & season == "Spring") > 0 ~ "Spring",
    sum(presence_absence == 1 & season == "Fall") > 0 ~ "Fall",
    TRUE ~ "None"
  )) %>%
  ungroup() #%>% 
  #intentionally add 0s
  #group_by(siteid, siteid_star) %>% 
  # complete(siteid_star, siteid, commonname, fill = list(color_category = NA)) %>% 
  # filter(commonname != "Not recorded")


## order based on site order
PA_filter_order <-PA_summarized %>% 
  mutate(siteid_star =  factor(siteid_star, levels = site_order_star)) %>%
  arrange(siteid_star)  %>% 
  #order based on species count
  mutate(commonname =  factor(commonname, levels = SpeciesList)) %>%
  arrange(commonname)  

## define colors for color category
colors <- c("Spring" = "#DDCC77", "Fall" = "#88CCEE", "Both" = "#117733",
            "None" = "white")

##GGPLOT TIME 
#adjust data because we don't want Not recorded species
fish_filtered_data <- PA_filter_order %>%
  filter(commonname != "Not recorded") %>%
  mutate(siteid_star = factor(siteid_star, levels = unique(PA_filter_order$siteid_star)))

n_species <- length(unique(fish_filtered_data$commonname))

heatmap <- fish_filtered_data %>% 
  ggplot(aes(commonname, siteid_star, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Fish Species", y = "SiteID", title = "2024 Fish Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Fish/Fish_PA_heatmap.jpg", 
           heatmap,
       width = 12,
       height = 8,
           dpi = 150)

```

```{r Simplified nMDS}

str(PA_filter)

Fish_PA_data_wide <- PA_filter %>% 
  #select variables for comparison
  select(siteid, siteid_star, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season,
         commonname, presence_absence) %>%
  #get rid of sites with Not recorded fish, therefore getting rid of the sites with no fish
  filter(commonname != "Not recorded") %>% 
  #Switch to wide format
  pivot_wider(names_from = commonname, values_from = presence_absence, values_fill = 0 ) %>%
  unite("SiteID_full", siteid, stationno, netreplicate, date,
        sep = "-",
        remove = FALSE ) #%>% 
  #filter(SiteID_full != "SC-VEN-3-2-2023-11-03")

Fish_meta_df <- Fish_PA_data_wide %>% 
  select(siteid, siteid_star, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season) 

Fish_matrix <- Fish_PA_data_wide %>% 
  select(-c(siteid, siteid_star, estuaryname, date, stationno, surveytype, netreplicate, region,
         estuaryclass, mpastatus, estuarytype, year, season)) %>% 
  column_to_rownames("SiteID_full")

#Permanvoa

Fish_perm <- adonis2(Fish_matrix ~ season,
                data = Fish_meta_df)
Fish_perm


#nmds

nmds.Fish <- metaMDS(Fish_matrix, distance = "bray", autotransform = TRUE)
#autotransform = TRUE for transformation, otherwise keep FALSE
nmds.Fish #stress = 0.112 6/29/25

plot(nmds.Fish)

#look for outliers
orditorp(nmds.Fish, "sites") 
#identify the outliers in order to drop from analysis = SC-BAT-1-3

# Stress Plot
stressplot(nmds.Fish, main = "Shepard plot")

# Extract scores

# Extract site scores (samples)
site.scores <- as.data.frame(scores(nmds.Fish, display = "sites"))
site.scores$SampleID <- rownames(site.scores)  # optional for labeling

# Extract species scores (species vectors)
species.scores <- as.data.frame(scores(nmds.Fish, display = "species"))
species.scores$Species <- rownames(species.scores)

# Add metadata to sites
site.scores <- cbind(site.scores, Fish_meta_df)

####extract signficant species
#this can be repeated for enironmental vectors
#replicae "vectors" with "factors"

# Fit species to the NMDS ordination
fish_envfit <- envfit(nmds.Fish, Fish_matrix, permutations = 999)

# Extract p-values
species_pvals <- as.data.frame(fish_envfit$vectors$pvals)
species_scores <- as.data.frame(scores(fish_envfit, display = "vectors"))
species_r <- as.data.frame(fish_envfit$vectors$r)

# Combine
species_scores$Species <- rownames(species_scores)
species_scores$pval <- species_pvals[, 1]
species_scores$r <- species_r[, 1]

# Keep only p < 0.01
species_sig <- species_scores %>%
  filter(pval < 0.01)

#Rescale NMDS vectors by r
species_sig <- species_sig %>%
  mutate(NMDS1_scaled = NMDS1 * r * 3,
         NMDS2_scaled = NMDS2 * r * 3)


# PLot

nMDS.plot.fish <- site.scores %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot() +
  # Sample points
  geom_point(data = site.scores, aes(x = NMDS1, y = NMDS2, 
                                     color = season,
                                     shape = siteid), 
             size = 3) +
  
  #Confidence ellipses
  stat_ellipse(data = site.scores,
               aes(x = NMDS1, y = NMDS2, color = season),
               type = "t", level = 0.95, linetype = "dashed", size = 1) +
  #type = "t" uses a multivariate t-distribution (better for small sample sizes).
  #level = 0.95 is for 95% confidence ellipses.
  #You can switch to type = "norm" if assumptions for t-distribution don’t hold.

  # Arrows scaled by r
  geom_segment(data = species_sig %>%
                 arrange(desc(r)) %>%   # Sort descending by r
                 slice_head(n = 5),      # Take top 5 rows,
               aes(x = 0, y = 0, xend = NMDS1_scaled, yend = NMDS2_scaled),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +

  #Species labels at arrow tips
  geom_text_repel(
    data = species_sig %>%
                 arrange(desc(r)) %>%   # Sort descending by r
                 slice_head(n = 5),
    aes(x = NMDS1_scaled, y = NMDS2_scaled, label = Species),
    size = 4, fontface = "bold", color = "black",
    nudge_y = 0.05, # small nudge to prevent overlap with arrow tip
    segment.size = 0.5,
    segment.color = "grey50"
  ) +
  
  #aesthetics 
  theme_bw() + 
  labs(title = "NMDS with Significant Species Vectors (Scaled by r)",
       x = "NMDS1", y = "NMDS2") +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10)) +
  coord_fixed() +  ## need aspect ratio of 1
  
  #add stress label
  annotate(geom = "label", x = 5, y = -1, size = 3,
           label = paste("Stress: ", round(nmds.Fish$stress, digits = 3))) +
  
  #add specifications for colors and shapes
  scale_shape_manual(name = "Site",
                     values=c(15, 16, 17, 18, 3, 4, 8)) +
  scale_color_manual(name = "Season",
                       values = c("#DDCC77", "#88CCEE"),
                     breaks = c("Spring", "Fall"))

nMDS.plot.fish


ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Fish/Fish_nMDS_species.jpg", 
           nMDS.plot.fish,
       width = 10,
       height = 6,
           dpi = 150)


```


## Metrics

```{r Species richness}

str(Fish_PA_data_wide)

#Richness by site

# First Calculate the total number of nets for each site
# Summarize the number of unique combinations of netreplicate and stationno for each site
netreplicate_data <- PA_filter %>% #we will use PA_filter because this includes all nets
  group_by(siteid_star, season) %>%
  summarize(total_replicates = n_distinct(paste(netreplicate, stationno)), .groups = 'drop')

#Calculate SR by site
Fish_PA_data_wide_SR_site <- Fish_PA_data_wide %>% 
  group_by(siteid, siteid_star, estuaryname, region,
         estuaryclass, mpastatus, estuarytype, year, season) %>% 
  summarise(across('California halibut':'Western mosquitofish', sum)) %>% 
  ungroup() %>% 
  #change numeric values to 1 so that it's PA
  mutate(across(c('California halibut':'Western mosquitofish'), ~ if_else(. > 0, 1, 0))) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'California halibut':'Western mosquitofish') %>% 
      rowSums()) %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  #add in 0 for species richness
  mutate(species_richness = ifelse(is.na(species_richness), 0, species_richness)) %>% 
  left_join(., netreplicate_data) %>% 
  #Arroyo Sequit Fall was not fished, so need to change the 0 back to NA
  mutate(species_richness = case_when(
    siteid == "SC-ASE" & season == "Fall" ~ NA,
    TRUE ~ species_richness
  ))


SR_site_plot <- Fish_PA_data_wide_SR_site %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid_star, y=species_richness, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Fish species richness") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) + 
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  geom_text(aes(label = total_replicates), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5)   # Show total replicates as label

SR_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Fish/Fish_SpeciesRichness_bysite.jpg", 
           SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Fish_SR_site_metric <- Fish_PA_data_wide_SR_site %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, species_richness, total_replicates) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(species_richness, 4)) %>% 
  #add metric name
  mutate(metric = "fish richness") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(species_richness),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 
  

```

```{r Size range}

str(Length_Data_clean)

#calculate the range of fish sizes

Fish_size_range <- Length_Data_clean %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  #select only fish
  filter(fish_or_invert == "fish") %>% 
  filter(year == "2024") %>% 
  group_by(siteid, siteid_star, season,
           estuaryname, estuarytype, year) %>% 
  summarise(fish_range = max(length_mm) - min(length_mm))

#lets plot a full boxplot of all the lengths 

Fish_size_boxplot <- Length_Data_clean %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x = siteid_star, y = length_mm, fill = season)) +
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), width = 0.6) +
  geom_jitter(aes(color = season, group = season),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  scale_color_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  labs(x = "Site", y = "Fish length (mm)") +
  scale_x_discrete(limits = (levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom"
  ) +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) 

Fish_size_boxplot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Fish/Fish_SizeRange_bysite.jpg", 
           Fish_size_boxplot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Fish_size_site_metric <- Fish_size_range %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, fish_range) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(fish_range, 4)) %>% 
  #add metric name
  mutate(metric = "size_range") %>% 
  #I like the other quartiles better
    mutate(
    quartile_2 = cut(
      fish_range,
      breaks = unique(quantile(fish_range, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  #new quartiles
    mutate(
    pct_rank = percent_rank(fish_range),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) #%>%
  #select(-pct_rank) %>%  # remove helper column
 


```

```{r tideater gobies}

#we want to make a metric for presence of tidewater gobies

str(PA_summarized) # this will tell you what season they were found

Goby_presence <- Fish_PA_data_wide_SR_site %>% 
  select(siteid:season, tidewater_goby = 'tidewater goby') %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  #add in 0 for species richness
  mutate(tidewater_goby = ifelse(is.na(tidewater_goby), 0, tidewater_goby)) %>% 
  left_join(., netreplicate_data) %>% 
  #Arroyo Sequit Fall was not fished, so need to change the 0 back to NA
  mutate(tidewater_goby = case_when(
    siteid == "SC-ASE" & season == "Fall" ~ NA,
    TRUE ~ tidewater_goby
  )) %>% 
  mutate(score = case_when(
    tidewater_goby == 1 ~ 1,
    tidewater_goby == 0 ~ 0
  )) %>% 
  mutate(metric = "tidewater_goby_presence")




```

#Macroalgae

```{r Data}

#we can look at the % of SAV or simply the presence 

#floating data
SOP_7_macro_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_macroalgae_floating")

Macro_Floating_df <- SOP_7_macro_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, surveyarea, estimatedcover,
         stationno, scientificname, commonname,
         status, region:estuarytype) %>% 
  #remove -88
  filter(estimatedcover != -88) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))


#Intertidal
SOP_7_intertidal_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_macroalgae_transect_meta")

Macro_Intertidal_df <- SOP_7_intertidal_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, transectlocation,
         transectreplicate, plotreplicate, non_algae_cover, total_algae_cover,
         region:estuarytype)  %>%
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
    #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))



```

## Figures

```{r Floating barplot}

Floating_plot <- Macro_Floating_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, estuarytype,
           year, region, season) %>% 
  summarise(Macro_avg = mean(estimatedcover, na.rm = TRUE),
            Macro_SD = sd(estimatedcover, na.rm = TRUE)) %>% 
  #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = rev(site_order))) %>% 
  #mutate(year = factor(year, levels = c("2023", "2024"))) %>% 
  ungroup() %>%  # Ungroup the data before completing
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid, year, season, fill = list(Macro_avg = NA, Macro_SD = NA))  %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid, y=Macro_avg, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(0.9)) +
  #prevent the error bars from going below 0
  geom_errorbar(aes(ymin= pmax(Macro_avg-Macro_SD, 0), ymax=Macro_avg+Macro_SD), 
                width=0.2, position= position_dodge(0.9)) +
  ylab("Avg. % Floating Macroalgae Cover") + 
  xlab("Site") +
  theme_bw() +
  #facet_wrap(~year) %>% 
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  coord_flip() 


Floating_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/SAV/Macro_floating.jpg", 
           Floating_plot,
       width = 10,
       height = 8,
           dpi = 150)

```

```{r stressors }
# a few stressor plots
temp_do <- read.csv("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/data-inventory/DO_temp_SMBay.csv")

do_plot <- ggplot(temp_do) +
  geom_bar(aes(x = siteid, y = num_events_DO), stat = "identity", fill = "darkred") +
  ylab("number of annual DO events < 3 mg/L") +
  xlab("estuary") +
  theme_bw() +
  facet_wrap(~season) +
  theme(axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size =24),
        axis.text.x = element_text(size = 18, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 18, angle = 45, hjust = 1),
        strip.text = element_text(size = 20)) 
  
 
do_plot

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figs_edit/DO_plot.jpg",
       do_plot, 
       height = 12,
       width = 12,
       dpi = 300)

temp_plot <- ggplot(temp_do) +
  geom_bar(aes(x = siteid, y = perc_temp_abv_25), stat = "identity", fill = "red") +
  ylab("average percent time above 25C") +
  xlab("estuary") +
  theme_bw() +
  ylim(c(0,100)) +
  facet_wrap(~season) +
   theme(axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size =18),
        axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 16, angle = 45, hjust = 1),
        strip.text = element_text(size = 18)) 
  
temp_plot

combined_stressor <- cowplot::plot_grid(do_plot, temp_plot, labels = c("A", "B"), label_size = 16, ncol = 1)
combined_stressor

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/figs_edit/temp_do_plot.jpg",
       combined_stressor, 
       height = 12,
       width = 12,
       dpi = 300)


#nonnative veg
nonnative_veg <- Veg_native_nonnative %>% filter(status == "invasive") %>% filter(siteid %in% c("SC-ASE", "SC-BSE", "SC-ZUM", "SC-MAL", "SC-TOP"))

#nonnative fish
Seine_nonnative <- Seine_Data_clean %>%
  filter(siteid %in% c("SC-ASE", "SC-BSE", "SC-ZUM", "SC-MAL", "SC-TOP"))%>%
  filter(method == "count") %>%
   mutate(sample_date = ymd(date)) %>% 
  #just keep the date, no time
  mutate(sample_date = date(sample_date)) %>% 
  #remove -88
  #filter(estimatedcover != -88) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  filter(status %in% c("non-native", "invasive")) %>%
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>%
  mutate(species_label = paste(scientificname, commonname, sep = " (") %>% 
                              paste0(")"))
  
  
invasive_fish_table <- Seine_nonnative %>%
  group_by(siteid, year, season) %>%
  summarize(
    invasive = paste(unique(scientificname[status == "invasive"]), collapse = ","),
    nonnative = paste(unique(scientificname[status == "non-native"]), collapse = ","),
    .groups = "drop")
  
library(knitr)
library(kableExtra)

invasive_fish_table %>%
  kable(format = "html",
        caption = "Non-native and invasive fish by estuary and season",
        escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1, bold = TRUE) %>%
  collapse_rows(columns = 1)
  
```

```{r Floating stacked}

Floating_stacked_plot <- Macro_Floating_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, estuaryclass, estuarytype,
           year, region, season) %>% 
  summarise(algae_avg = mean(estimatedcover, na.rm = TRUE)) %>% 
  mutate(nonalgae_avg = 100-algae_avg) %>% 
  pivot_longer(cols = c(algae_avg, nonalgae_avg), names_to = "Cover") %>% 
  #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = rev(site_order))) %>% 
  mutate(Cover = factor(Cover, levels = c("nonalgae_avg", "algae_avg"))) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid, y=value, fill = Cover)) + 
  geom_bar(stat = "identity", color = "black", position = "stack") +
  ylab("Avg. % Floating Macroalgae Cover") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Cover Type",
                    values = c("#999999", "#009E73"),
                    breaks=c("nonalgae_avg", "algae_avg"),
                    labels=c("non-algae", "algae")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  facet_wrap(~season) +
  coord_flip() 


Floating_stacked_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/SAV/Macro_floating_stacked.jpg", 
           Floating_stacked_plot,
       width = 10,
       height = 8,
           dpi = 150)

```


```{r intertidal}

#we want to look at the %alge to $non-algae

Intertidal_plot <- Macro_Intertidal_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, estuaryclass, estuarytype,
           year, region, season) %>% 
  summarise(algae_avg = round(mean(total_algae_cover, na.rm = TRUE), 0),
            nonalgae_avg = round(mean(non_algae_cover, na.rm = TRUE), 0)) %>% 
  pivot_longer(cols = c(algae_avg, nonalgae_avg), names_to = "Cover") %>% 
  #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = rev(site_order))) %>% 
  mutate(Cover = factor(Cover, levels = c("nonalgae_avg", "algae_avg"))) %>% 
  ggplot(aes(x=siteid, y=value, fill = Cover)) + 
  geom_bar(stat = "identity", color = "black", position = "stack") +
  ylab("Avg. % Intertidal Macroalgae Cover") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Cover Type",
                    values = c("#999999", "#009E73"),
                    breaks=c("nonalgae_avg", "algae_avg"),
                    labels=c("non-algae", "algae")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  facet_wrap(~season) +
  coord_flip() 


Intertidal_plot

#transects were not completed at the majority of sites

```

## Metrics

```{r % Floating}

Macro_Floating_metric <- Macro_Floating_df %>% 
  filter(year == 2024) %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  group_by(siteid, estuaryname, estuaryclass, estuarytype,
           year, region, season) %>% 
  summarise(algae_avg = mean(estimatedcover, na.rm = TRUE)) %>% 
  mutate(nonalgae_avg = 100-algae_avg) %>% 
  ungroup() %>% 
  group_by(season) %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      nonalgae_avg,
      breaks = unique(quantile(nonalgae_avg, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  mutate(metric = "nonalgae_cover") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(nonalgae_avg),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 
#both quartile methods provide the same results, only breaks data into 1 and 2

```

# SAV

```{r Data}

#we record the presence of SAV in two locations
#Discrete WQ
# Fish

#we will pull the data from discrete WQ and fish and then indicate if there were any records of SAV

SOP_2_WQ_SAV_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_waterquality_metadata")

SAV_WQ_df <- SOP_2_WQ_SAV_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, profile, sav_presence,
         speciesname, region:estuarytype) %>% 
  #remove -88
  #filter(estimatedcover != -88) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))


SOP_9_fish_SAV_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_fish_sample_metadata")

Fish_WQ_df <- SOP_9_fish_SAV_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, netreplicate, savpresent,
         region:estuarytype) %>% 
  #remove -88
  #filter(estimatedcover != -88) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))


```

##Metrics

```{r Presence}

#determine if SAV is ever present within the estuary

SAV_fish_presence <- Fish_WQ_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, season, estuarytype) %>% 
  count(savpresent) %>%
  left_join(Site_df, .)

SAV_WQ_presence <- SAV_WQ_df %>% 
  filter(year == 2024) %>% 
  group_by(siteid, estuaryname, season, estuarytype) %>% 
  count(sav_presence) %>% 
  #we only care about if a site had SAV
  filter(sav_presence == "present") 

SAV_both <- SAV_fish_presence %>% 
  #overwrite Top bc there was SAV present in the Fall, based on WQ data
  mutate(savpresent = case_when(
    siteid == "SC-TOP" & season == "Fall" ~ "Yes",
    TRUE ~ savpresent
  ))


#create a full table

SAV_presence_metric <- SAV_both %>% 
  mutate(score = case_when(
    savpresent == "Yes" ~ 1,
    savpresent == "No" ~ 0
  )) %>% 
  mutate(metric = "SAV_presence") %>% 
  group_by(siteid, estuaryname, season, estuarytype) %>% 
  filter(if (any(savpresent == "Yes", na.rm = TRUE)) savpresent == "Yes" else TRUE) %>% 
  ungroup()
  

```

## Figures

```{r heatmap}

n_seasons <- SAV_presence_metric %>% pull(season) %>% n_distinct()

SAV_heatmap <- SAV_presence_metric %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  mutate(siteid_star = factor(siteid_star, levels = rev(site_order_star))) %>% 
  ggplot(aes(season, siteid_star, fill = savpresent))+
  geom_tile(colour = "black", linewidth = 0.5) +
  scale_fill_manual(name = "SAV Present",
                    values = c("#999999", "#009E73"),
                    breaks=c("No", "Yes"),
                    na.value = "white") +
  #coord_equal(expand = 0) +
  #scale_y_discrete(limits = (levels(SAV_presence_metric$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 14), # bold common names
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16, face = "bold"),
        axis.title.x = element_text(size = 16, face = "bold"),
        legend.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 14),
        title = element_text(size = 16, face = "bold")
        
  )+
  labs(x = "Season", y = "SiteID") 

## print it
SAV_heatmap

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/SAV/SAV_presence.jpg", 
           SAV_heatmap,
       width = 10,
       height = 8,
           dpi = 150)



```

# Sediment

```{r Data}

##### GS

#pull in grainsize
SOP_3_gs_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_sedgrainsize_data")
str(SOP_3_gs_data)

SOP_3_gs_data_clean <- SOP_3_gs_data %>% 
  # mutate date
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  # add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC")) %>% 
  #average per station
  group_by(siteid, estuaryname, stationno, date, year, season, region,
           estuarytype, mpastatus, samplelocation, wentworth_class) %>%
  summarise(gs_avg = mean(value)) %>% 
  #create new classes
  mutate(condensed_class = case_when(
    str_detect(wentworth_class, "sand") ~ "sand",
    str_detect(wentworth_class, "silt") ~ "silt",
    str_detect(wentworth_class, "clay") ~ "clay",
    TRUE ~ wentworth_class
  )) %>% 
  ungroup() %>% 
  #pool into groups by condensed class
  group_by(siteid, estuaryname, stationno, date, year, season, region,
           estuarytype, mpastatus, samplelocation, condensed_class) %>%
  summarise(gs_pooled = sum(gs_avg)) %>% 
  #pivot wider
  pivot_wider(names_from = condensed_class, values_from = gs_pooled, values_fill = 0) %>% 
  #calculate percent sand
  mutate(percent_sand = (sand/(sand + silt + clay)*100)) %>% 
  mutate(percent_mud = ((silt + clay)/(sand + silt + clay)*100)) %>% 
  #relabel sand or mud
  mutate(SedType=if_else(percent_sand>=60, "Sand", "Mud"))


#### Nutrients

SOP3_Nutrients_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_sedchem_data")

SOP3_Nutrients_data_clean <- SOP3_Nutrients_data %>% 
  # mutate date
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  # add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))

SOP3_Nutrients_data_aggreate <- SOP3_Nutrients_data_clean %>%
  #select on sediment
  filter(matrix == "sediment") %>% 
  #get rid of -88
  filter(results != -88) %>% 
  #convert to mg_g
  mutate(Result_mgkg = results*10000,
         Result_mg_g = Result_mgkg/1000) %>% 
  # Summarize data by siteid, stationno, date, Year, and Season, analyte.
  group_by(siteid, stationno, date, year, season, estuarytype, analytename) %>%
  summarise(result_avg = mean(Result_mg_g)) %>% 
  #pivot wide
  pivot_wider(names_from = analytename, values_from = result_avg) %>% 
  rename(TOC = "Sediment Particulate Organic Carbon", 
         TN = "Sediment Particulate Nitrogen") %>% 
  ungroup()


#combine with nuts

SOP3_Nuts_gs_aggregate <- SOP3_Nutrients_data_aggreate %>% 
  left_join(., SOP_3_gs_data_clean) %>% 
pivot_longer(cols = c(TOC, TN), names_to = "parameter")

```

## Figures

```{r Boxplot}

Nuts_boxplot <- SOP3_Nuts_gs_aggregate %>% 
  filter(year == 2024) %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>%
  mutate(siteid_star = factor(siteid_star, levels = rev((site_order_star)))) %>%
  ggplot(aes(x = siteid_star, y = value, fill = season)) +
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), width = 0.6) +
  geom_jitter(aes(color = season, group = season),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 2, alpha = 0.7) +
  scale_fill_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  scale_color_manual(values = c("Spring" = "#DDCC77", "Fall" = "#88CCEE")) +
  facet_wrap(~parameter, scales="free") +
  labs(x = "Site", y = "mg/g") +
  #scale_x_discrete(limits = (levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(size = 14, face = "bold")
  ) +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  coord_flip()

Nuts_boxplot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Sediment/Nuts-boxplot.jpg", 
           Nuts_boxplot,
       width = 10,
       height = 8,
           dpi = 150)


```

```{r stacked barplot}

Sediment_stacked <- SOP3_Nuts_gs_aggregate %>% 
  filter(year == 2024) %>% 
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>%
  mutate(siteid_star = factor(siteid_star, levels = rev((site_order_star)))) %>%
  select(siteid:samplelocation, percent_mud, percent_sand, siteid_star) %>% 
  pivot_longer(cols = c(percent_mud, percent_sand), names_to = "type") %>% 
  unique() %>% 
  group_by(siteid_star, season, type) %>% 
  summarise(avg_perc = mean(value)) %>% 
  ggplot(aes(x=siteid_star, y= avg_perc, fill= type)) +
  geom_col(position= "fill") +  # Stacks bars for each estuaryname so that they're proportional to their sum, creating a 100% stacked bar chart
  # height of each bar represents 100%, relative heights of the segments within the bar show the proportion of each wentworth_class.
  # geom_text(
  #   data = summary_wentworth,  # Use summarized data for labels
  #   aes(x = estuaryname, y = 1, label = total_samples),  # Position at the top of the bar
  #   inherit.aes = FALSE,  # Ensure it doesn't inherit aesthetics from ggplot
  #   vjust = -0.5  # Adjust label position
  # ) +
  theme_bw() +
  scale_fill_manual(name = "Sediment type",
                    values=c("brown", "#F0E442"),
                    labels = c("mud", "sand")) +
  facet_wrap(~season, scales="free") +
  xlab("Site") +
  ylab("Proportion") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    strip.text = element_text(size = 14, face = "bold")
  ) +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  coord_flip()

Sediment_stacked

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Sediment/Grainsize_stacked.jpg", 
           Sediment_stacked,
       width = 10,
       height = 8,
           dpi = 150)

```

##Metrics

```{r avg. TN}

Avg_TN_metric <- SOP3_Nuts_gs_aggregate %>% 
  filter(year == 2024) %>% 
  filter(parameter == "TN") %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  group_by(siteid, estuaryname, estuarytype,
           year, region, season, parameter) %>% 
  summarise(TN_avg = mean(value, na.rm = TRUE)) %>% 
  filter(!is.na(TN_avg)) %>% 
  ungroup() %>% 
  # pivot_wider(names_from = parameter, values_from = nuts_avg) %>% 
  group_by(season) %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      TN_avg,
      breaks = unique(quantile(TN_avg, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  mutate(metric = "TN_mg/g") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(TN_avg),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 


```

```{r avg-TOC}

Avg_TOC_metric <- SOP3_Nuts_gs_aggregate %>% 
  filter(year == 2024) %>% 
  filter(parameter == "TOC") %>% 
  #add in missing istes
  left_join(Site_df, .) %>% 
  group_by(siteid, estuaryname, estuarytype,
           year, region, season, parameter) %>% 
  summarise(TOC_avg = mean(value, na.rm = TRUE)) %>% 
  filter(!is.na(TOC_avg)) %>% 
  ungroup() %>% 
  # pivot_wider(names_from = parameter, values_from = nuts_avg) %>% 
  group_by(season) %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      TOC_avg,
      breaks = unique(quantile(TOC_avg, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  mutate(metric = "TOC_mg/g") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(TOC_avg),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 

```

# Vegetation

```{r Data}

# Veg metatdata table
SOP_11_vegsample_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_vegetation_sample_metadata")

# Veg cover table 
SOP_11_vegcover_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_vegetativecover_data")

#####Veg cover

Vegcover_df <- SOP_11_vegcover_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, stationno, date, method,
         habitat, transectreplicate, plotreplicate, covertype, live_dead,
         scientificname, commonname, status, estimatedcover, tallestplantheight_cm,
        region:estuarytype) %>% 
# add season
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))



####### veg sample data

Vegsample_Data_clean <- SOP_11_vegsample_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, stationno, date, method,
         habitat, transectreplicate, plotreplicate, vegetated_cover,
         non_vegetated_cover,
        region:estuarytype) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))



```

## Community

```{r Heatmap}

PA_veg <- Vegcover_df %>%
  #add star so can flag in graphs where sites are open
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  filter(year == 2024) %>% 
  #remove Not recorded
  filter(scientificname != "Not recorded") %>% 
  #keep only vegetation cover class
  filter(covertype == "vegetation") %>% 
  #filter out unknowns
  filter(!grepl("unknown", commonname, ignore.case = TRUE)) %>% 
  mutate(presence_absence = ifelse(estimatedcover != 0, 1, 0)) %>% 
  #only filter SM Bay sites bc other sites sampled Fall only, and only in marsh
  filter(siteid %in% c("SC-MAL", "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))


## find order of PA from most to least for ordering of heat map
SpeciesOrder <- PA_veg %>%
  filter(presence_absence == 1) %>% 
  select(scientificname, siteid) %>% 
  distinct() %>% 
  group_by(scientificname) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

## get the ordered list of species by their count
SpeciesList <- SpeciesOrder %>% pull(scientificname)

## summarize df and create color category for PA heat map. will add new color if present in both years 
PA_summarized <- PA_veg %>%
  group_by(siteid, siteid_star, scientificname) %>%
  summarise(color_category = case_when(
    sum(presence_absence == 1 & season == "Fall") > 0 & 
      sum(presence_absence == 1 & season == "Spring") > 0 ~ "Both",
    sum(presence_absence == 1 & season == "Spring") > 0 ~ "Spring",
    sum(presence_absence == 1 & season == "Fall") > 0 ~ "Fall",
    TRUE ~ "None"
  )) %>%
  ungroup() #

## order based on site order
PA_filter_order <-PA_summarized %>% 
  mutate(siteid_star =  factor(siteid_star, levels = site_order)) %>%
  arrange(siteid_star)  %>% 
  #order based on species count
  mutate(scientificname =  factor(scientificname, levels = SpeciesList)) %>%
  arrange(scientificname)  

## define colors for color category
colors <- c("Spring" = "#DDCC77", "Fall" = "#88CCEE", "Both" = "#117733",
            "None" = "white")

##GGPLOT

n_species <- length(unique(PA_filter_order$scientificname))

heatmap <- PA_filter_order %>% 
  ggplot(aes(scientificname, siteid_star, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Plant Species", y = "SiteID", title = "2024 Plant Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Veg/Veg_PA_heatmap.jpg", 
           heatmap,
       width = 16,
       height = 6,
           dpi = 150)

```

## Metrics

```{r Richness}

str(PA_veg)

Veg_PA_data_wide_SR <- PA_veg %>% 
  #select variables for comparison
  select(siteid, siteid_star, estuaryname, date, stationno, transectreplicate, plotreplicate,
         method, habitat, covertype, live_dead, region, estuarytype, year, season,
         scientificname, presence_absence) %>% 
  #Switch to wide format
  pivot_wider(names_from = scientificname, values_from = presence_absence, values_fill = 0 ) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Salicornia pacifica':'Lemna minor') %>% 
      rowSums())

######

#Richness by site

#Calculate SR by site
Veg_PA_data_wide_SR_site <- Veg_PA_data_wide_SR %>% 
  group_by(siteid, siteid_star, estuaryname, region, estuarytype, year, season) %>% 
  summarise(across('Salicornia pacifica':'Lemna minor', sum)) %>% 
  ungroup() %>% 
  #change numeric values to 1 so that it's PA
  mutate(across(c('Salicornia pacifica':'Lemna minor'), ~ if_else(. > 0, 1, 0))) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Salicornia pacifica':'Lemna minor') %>% 
      rowSums())

Veg_SR_site_plot <- Veg_PA_data_wide_SR_site %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(species_richness = NA))  %>% 
  filter(siteid_star != "SC-GOL*",
         siteid_star != "SC-VEN*",
         siteid_star != "SC-NEW",
         siteid_star != "SC-BAT") %>% 
  ggplot(aes(x=siteid_star, y=species_richness, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Plant species richness") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) + 
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

Veg_SR_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Veg/Veg_SpeciesRichness_bysite.jpg", 
           Veg_SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Veg_SR_site_metric <- Veg_PA_data_wide_SR_site %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, species_richness) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(species_richness, 4)) %>% 
  #add metric name
  mutate(metric = "plant richness") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(species_richness),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 


```

```{r % native}

#want to look at % of total species that was invasive 

Veg_native_nonnative <- Vegcover_df %>% 
  #remove Not recorded
  filter(scientificname != "Not recorded") %>% 
  filter(year == 2024) %>% 
  #keep only vegetation cover class
  filter(covertype == "vegetation") %>% 
  #filter out unknowns
  filter(!grepl("unknown", commonname, ignore.case = TRUE)) %>% 
  filter(status != "Not recorded") %>% 
  #change invasive to non-native because we'll treat them the same
  mutate(status = case_when(
    status == "non-native" ~ "invasive",
    status == "native and non-native" ~ "native",
    TRUE ~ status
  )) %>% 
  #add up the total species for native and invasive
  group_by(siteid, estuaryname, status, year, season, estuarytype) %>% 
  summarise(total_species = n_distinct(scientificname), .groups = "drop") %>% 
  group_by(siteid, estuaryname, year, estuarytype, season) %>% 
  mutate(proportion = total_species / sum(total_species))


Veg_Native_site_plot <- Veg_native_nonnative %>% 
  #add star so can flag in graphs
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  #added completet to fill in missing combinations to reserve bra position
  #complete(siteid_star, season, status, fill = list(proportion = NA))  %>% 
  filter(siteid_star != "SC-GOL*",
         siteid_star != "SC-VEN*",
         siteid_star != "SC-NEW",
         siteid_star != "SC-BAT") %>% 
  ggplot(aes(x=siteid_star, y=proportion, fill = status)) + 
  geom_bar(stat = "identity", color = "black") +
  ylab("% native plants") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~season) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Status",
                    values = c("red", "#88CCEE")) + #TYPE
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

Veg_Native_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Veg/Veg_perc_native.jpg", 
           Veg_Native_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Veg_native_metric <- Veg_native_nonnative %>% 
  #add star so can flag in graphs
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  filter(siteid_star != "SC-GOL*",
         siteid_star != "SC-VEN*",
         siteid_star != "SC-NEW",
         siteid_star != "SC-BAT") %>% 
  #select on native for netric
  filter(status == "native") %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, proportion) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(proportion, 4)) %>% 
  #add metric name
  mutate(metric = "percent native") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(proportion),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 


```

# CRAM

```{r Data}

#call in CRAM data
#downloaded on June 20, 2025

CRAM_data <- read.csv("/Users/janw/Desktop/projects/Bight-Estuary-2023/Data/ecoatlas_wetlandcondition-6-20-25.csv")

CRAM_data_clean <- CRAM_data %>% 
  select(vyear, aaname, visitdate, county, ecoregion, indexscore:biotic_structure) %>% 
  filter(ecoregion == "south coast" |
           county == "Santa Barbara") %>% 
  #crosswalk names
  mutate(estuaryname = case_when(
   str_detect(aaname, "Big Sycamore") ~ "Big Sycamore Canyon",
   str_detect(aaname, "Zuma") ~ "Zuma Lagoon",
   str_detect(aaname, "Topanga") ~ "Topanga Lagoon",
   str_detect(aaname, "Batiquitos") ~ "Batiquitos Lagoon",
   str_detect(aaname, "SC-VEN") ~ "Ventura River",
   str_detect(aaname, "Ventura") ~ "Ventura River",
   str_detect(aaname, "SC-GOL") ~ "Goleta Slough",
   str_detect(aaname, "Goleta Slough") ~ "Goleta Slough",
   str_detect(aaname, "SC-MAL") ~ "Malibu Lagoon",
   str_detect(aaname, "Malibu") ~ "Malibu Lagoon",
   str_detect(aaname, "NWCA21-CA-10156-San Diego Bay") ~ "Sweetwater Marsh",
   str_detect(aaname, "Newport") ~ "Newport Bay",
   str_detect(aaname, "Mugu") ~ "Point Mugu",
   str_detect(aaname, "Tijuana") ~ "Tijuana River Estuary",
   str_detect(aaname, "Los Pen") ~ "Los Penasquitos Lagoon",
   str_detect(aaname, "Salt Pond") ~ "San Diego Bay: South salt ponds",
   str_detect(aaname, "Mission Bay") ~ "Mission Bay: Kendall-Frost Reserve",
   str_detect(aaname, "San Dieguito") ~ "San Dieguito Lagoon",
   str_detect(aaname, "San Elijo") ~ "San Elijo Lagoon",
   str_detect(aaname, "Agua Hedionda") ~ "Agua Hedionda",
   str_detect(aaname, "Talbert") ~ "Huntington Beach Wetlands",
   str_detect(aaname, "Brookhurst") ~ "Huntington Beach Wetlands",
   str_detect(aaname, "Magnolia") ~ "Huntington Beach Wetlands",
   str_detect(aaname, "Seal") ~ "Seal Beach",
   str_detect(aaname, "Arroyo Sequit") ~ "Arroyo Sequit",
   str_detect(aaname, "Carpenteria Salt Marsh") ~ "Carpinteria Estuary",
   str_detect(aaname, "Carpenteria Marsh") ~ "Carpinteria Estuary",
   str_detect(aaname, "Upper Devereux") ~ "Devereaux Slough",
   #str_detect(aaname, "Bolsa Chica") ~ "Bolsa Chica", #the CRAM area is not the same as the site, so ignore for now
   TRUE ~ "Not recorded"
  )) %>% 
  left_join(., Site_df %>% 
              select(siteid, estuaryname)) %>% 
  #only select estaruies we want
  filter(!is.na(siteid))
  
  top_dates <- CRAM_data_clean %>%
  mutate(visitdate = mdy(visitdate)) %>%
  group_by(siteid) %>%
  arrange(desc(vyear)) %>%
  distinct(siteid, vyear, .keep_all = FALSE) %>%
  slice_head(n = 2) %>%
  ungroup()
  
  
CRAM_reduced <- CRAM_data_clean %>% 
  semi_join(., top_dates, by = c("siteid", "vyear")) %>%   # <-- Keep only rows with the top 2 dates
    group_by(siteid) %>%
  mutate(
    year_label = case_when(
      vyear == max(vyear) ~ "year 2",
      vyear == min(vyear) ~ "year 1"))

```

## Figures

```{r Avg - subindices}

CRAM_index <- CRAM_reduced %>% 
  group_by(siteid, vyear, year_label) %>% 
  summarise(index_avg = mean(indexscore),
            buffer_avg = mean(buffer_landscape_cont),
            hydrology_avg = mean(hydrology),
            physical_structure_avg = mean(physical_structure),
            biotic_structure_avg = mean(biotic_structure)) %>% 
  pivot_longer(cols = c(index_avg:biotic_structure_avg), names_to = "index")

index_labels <- c(
  buffer_avg = "Buffer",
  hydrology_avg = "Hydrology",
  physical_structure_avg = "Physcial Structure",
  biotic_structure_avg = "Biotic Structure",
  index_avg = "Index")


CRAM_plot_subindex <- CRAM_index %>% 
  filter(index != "index_avg") %>% 
  ungroup() %>% 
  complete(siteid, year_label, index, fill = list(value = NA, vyear = NA)) %>% 
  mutate(siteid = factor(siteid, levels = site_order)) %>% 
  mutate(year_label = factor(year_label, levels = c("year 1", "year 2"))) %>% 
  ggplot(aes(x=siteid, y=value, fill = year_label)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Score") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~index,
             labeller=labeller(index = index_labels)) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Survey",
                    values = c("#DDCC77", "#88CCEE"),
                    labels = c("Prior", "Recent")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) #+
  # geom_text(aes(label = vyear),
  #           position = position_dodge(width = 0.9),
  #           vjust = -0.5, size = 3.5,
  #           na.rm = TRUE)   # Show total replicates as label

CRAM_plot_subindex

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/CRAM/CRAM_subindices.jpg", 
           CRAM_plot_subindex,
       width = 10,
       height = 8,
           dpi = 150)


```

```{r Avg - Index}

CRAM_plot <- CRAM_index %>% 
  filter(index == "index_avg") %>% 
  ungroup() %>% 
  complete(siteid, year_label, index, fill = list(value = NA, vyear = NA)) %>% 
  mutate(siteid = factor(siteid, levels = site_order)) %>% 
  mutate(year_label = factor(year_label, levels = c("year 1", "year 2"))) %>% 
  ggplot(aes(x=siteid, y=value, fill = year_label)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Score") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~index,
             labeller=labeller(index = index_labels)) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Survey",
                    values = c("#DDCC77", "#88CCEE"),
                    labels = c("Prior", "Recent")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) #+
  # geom_text(aes(label = vyear),
  #           position = position_dodge(width = 0.9),
  #           vjust = -0.5, size = 3.5,
  #           na.rm = TRUE)   # Show total replicates as label

CRAM_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/CRAM/CRAM_index.jpg", 
           CRAM_plot,
       width = 10,
       height = 8,
           dpi = 150)


```

## Metrics

```{r index}

CRAM_index_metric <- CRAM_index %>%
  filter(year_label == "year 2") %>% 
  #add in missing istes
  left_join(Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype), .) %>% 
  group_by(index) %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      value,
      breaks = unique(quantile(value, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  group_by(index) %>%
  mutate(
    pct_rank = percent_rank(value),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) %>% 
  mutate(metric = case_when(
    str_detect(index, "biotic") ~ "CRAM_biotic",
    str_detect(index, "buffer") ~ "CRAM_buffer",
    str_detect(index, "hydrology") ~ "CRAM_hydrology",
    str_detect(index, "index") ~ "CRAM_index",
    str_detect(index, "physical") ~ "CRAM_physical"
  ))

```

#Trash

```{r Data}

SOP15_trash_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_trashtimesearchtally")

Trash_df <- SOP15_trash_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, habitat, debriscategory,
         debrisitem, resulttotal, resulttotaltext, region:estuarytype) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))
  

```

## Figures

```{r total}

Trash_df_total <- Trash_df %>% 
  filter(year == 2024) %>% 
  #get rid of -88 and NA
  filter(resulttotal != -88) %>% 
  filter(!is.na(resulttotal)) %>% 
  #avg trash within a 10min search
  group_by(siteid, estuaryname, stationno, season, region, estuarytype) %>% 
  summarise(total_trash = sum(resulttotal)) %>% 
  ungroup() %>% 
  group_by(siteid, estuaryname, season, region, estuarytype) %>% 
  summarise(total_trash_avg = mean(total_trash),
            total_trash_sd = sd(total_trash))

Trash_total_plot <- Trash_df_total %>% 
  #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = rev(site_order))) %>% 
  ungroup() %>% 
  ggplot(aes(x=siteid, y=total_trash_avg, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(0.9)) +
  #prevent the error bars from going below 0
  geom_errorbar(aes(ymin= pmax(total_trash_avg-total_trash_sd, 0), ymax=total_trash_avg+total_trash_sd), 
                width=0.2, position= position_dodge(0.9)) +
  ylab("Avg. trash total in 10min search") + 
  xlab("Site") +
  theme_bw() +
  #facet_wrap(~year) %>% 
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#88CCEE")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) +
  scale_x_discrete(limits = (levels(Trash_df$siteid)), drop = FALSE) +
  coord_flip() 


Trash_total_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Trash/Trash_total.jpg", 
           Trash_total_plot,
       width = 10,
       height = 8,
           dpi = 150)
  
```

```{r % plastic}

Trash_df_plastic <- Trash_df %>% 
  filter(year == 2024) %>% 
  #get rid of -88 and NA
  filter(resulttotal != -88) %>% 
  filter(!is.na(resulttotal)) %>% 
  #avg trash within a 10min search
  group_by(siteid, estuaryname, stationno, debriscategory, season, region, estuarytype) %>% 
  summarise(total_category = sum(resulttotal)) %>% 
  ungroup() %>% 
  group_by(siteid, estuaryname, debriscategory, season, region, estuarytype) %>% 
  summarise(total_category_avg = mean(total_category),
            total_category_sd = sd(total_category)) %>% 
  group_by(siteid, estuaryname, season, region, estuarytype) %>% 
  mutate(proportion = total_category_avg / sum(total_category_avg))

  

Perc_plastic_plot <- Trash_df_plastic %>% 
    #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = (site_order))) %>% 
  ggplot(aes(x=siteid, y=proportion, fill = debriscategory)) + 
  geom_bar(stat = "identity", color = "black") +
  ylab("% plastic") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~season) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Trash Type",
                    values = c("#DDCC77", "#88CCEE")) + #TYPE
  scale_x_discrete(limits = (levels(Trash_df$siteid)), drop = FALSE) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

Perc_plastic_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Trash/Trash_plastic_perc.jpg", 
           Perc_plastic_plot,
       width = 10,
       height = 8,
           dpi = 150)


```

```{r Debris items}

Trash_df_debris <- Trash_df %>% 
  filter(year == 2024) %>% 
  #get rid of -88 and NA
  filter(resulttotal != -88) %>% 
  filter(!is.na(resulttotal)) %>% 
  #avg trash within a 10min search
  group_by(siteid, estuaryname, stationno, debriscategory, debrisitem, season, region, estuarytype) %>% 
  summarise(total_trash = sum(resulttotal)) %>% 
  group_by(siteid, estuaryname, debriscategory, debrisitem, season, region, estuarytype) %>% 
  summarise(mean_trash = mean(total_trash)) %>% 
  group_by(siteid, estuaryname, season, region, estuarytype) %>% 
  mutate(proportion = mean_trash / sum(mean_trash))

debris_plot <- Trash_df_debris %>% 
    #flipped the order because we are going to have a coord flip later
  mutate(siteid = factor(siteid, levels = (site_order))) %>% 
  ggplot(aes(x=siteid, y=proportion, fill = debrisitem)) + 
  geom_bar(stat = "identity", color = "black") +
  ylab("% of total trash") + 
  xlab("Site") +
  theme_bw() +
  facet_wrap(~season) +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  # scale_fill_manual(name = "Trash Type",
  #                   values = c("#DDCC77", "#88CCEE")) + #TYPE
  scale_x_discrete(limits = (levels(Trash_df$siteid)), drop = FALSE) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

debris_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Trash/Perc_total_trash.jpg", 
           debris_plot,
       width = 10,
       height = 8,
           dpi = 150)



```

## Metrics

```{r total}

Trash_total_metric <- Trash_df_total %>%
  left_join(Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype), .) %>% 
  unique() %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      total_trash_avg,
      breaks = unique(quantile(total_trash_avg, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  mutate(
    pct_rank = percent_rank(total_trash_avg),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) %>% 
  mutate(metric = "total_trash")


```

```{r % plastic}

Trash_plastic_metric <- Trash_df_plastic %>%
  left_join(Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype), .) %>% 
  unique() %>% 
  filter(debriscategory == "Plastic" |
           is.na(debriscategory)) %>% 
#cut(..., breaks = quantile(...)) respects value distributions.
#unique() ensures no duplicate breakpoints (which can happen if you have tied values).
#include.lowest = TRUE puts the lowest value in the first bin.
#Labels 1–4 give you 4 quartiles.
  mutate(
    quartile = cut(
      proportion,
      breaks = unique(quantile(proportion, probs = seq(0, 1, 0.25), na.rm = TRUE)),
      include.lowest = TRUE,
      labels = FALSE
    )
  ) %>% 
  ungroup() %>%
  mutate(
    pct_rank = percent_rank(proportion),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) %>% 
  mutate(metric = "percent_plastic")

```

#Birds

```{r Data}

#call in flat files

bird_count_data <- read.csv("/Users/janw/Desktop/projects/Prop50-SMB/Data/Bird_counts_2024.csv")

bird_list <- read.csv("/Users/janw/Desktop/projects/Prop50-SMB/Data/bird_abbreviations.csv") 

bird_count_clean <- bird_count_data %>% 
  mutate(date = mdy(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>% 
  #add other site info
  left_join(., Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype) %>% 
              unique()) %>% 
  #change abundnace NA to 0
  mutate(abundance = replace_na(abundance, 0))

#add full bird names to abbreviations

bird_count_df <- bird_count_clean %>% 
  left_join(., bird_list %>% 
              select(birds = Abbreviation, commonname = Name))


```

## Community

```{r Heatmap}

PA_birds <- bird_count_df %>%
  mutate(presence_absence = ifelse(abundance != 0, 1, 0)) %>% 
  #malibu was sampled twice in spring 2024 - select commmon month
  filter(month != 5)


## find order of PA from most to least for ordering of heat map
SpeciesOrder <- PA_birds %>%
  filter(presence_absence == 1) %>% 
  select(commonname, siteid) %>% 
  distinct() %>% 
  group_by(commonname) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

## get the ordered list of species by their count
SpeciesList <- SpeciesOrder %>% pull(commonname)

## summarize df and create color category for PA heat map. will add new color if present in both years 
PA_summarized <- PA_birds %>%
  group_by(siteid, siteid_star, commonname) %>%
  summarise(color_category = case_when(
    sum(presence_absence == 1 & season == "Fall") > 0 & 
      sum(presence_absence == 1 & season == "Spring") > 0 ~ "Both",
    sum(presence_absence == 1 & season == "Spring") > 0 ~ "Spring",
    sum(presence_absence == 1 & season == "Fall") > 0 ~ "Fall",
    TRUE ~ "None"
  )) %>%
  ungroup() #

setdiff(unique(PA_summarized$commonname), SpeciesList) 
#more species are showing up then in the list because they are 0's
#we should filter them out

## order based on site order
PA_filter_order <-PA_summarized %>% 
  mutate(siteid_star =  factor(siteid_star, levels = site_order_star)) %>%
  arrange(siteid_star)  %>% 
  #order based on species count
  filter(commonname %in% SpeciesList) %>% 
  mutate(commonname =  factor(commonname, levels = SpeciesList)) %>%
  arrange(commonname)  

## define colors for color category
colors <- c("Spring" = "#DDCC77", "Fall" = "#88CCEE", "Both" = "#117733",
            "None" = "white")

##GGPLOT

n_species <- length(unique(PA_filter_order$commonname))


heatmap_birds <- PA_filter_order %>% 
  ggplot(aes(commonname, siteid_star, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Bird Species", y = "SiteID", title = "2024 Bird Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap_birds


ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Birds/Bird_PA_heatmap.jpg", 
           heatmap_birds,
       width = 16,
       height = 6,
           dpi = 150)



```

## Metrics

```{r species richness}

str(PA_birds)

Bird_PA_data_wide_SR <- PA_birds %>% 
  select(siteid, siteid_star, date, stationno, year, season, estuaryname, 
         estuarytype, commonname, presence_absence) %>% 
  #Switch to wide format
  pivot_wider(names_from = commonname, values_from = presence_absence, values_fill = 0 ) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Horned Grebe':'Hermit Thrush') %>% 
      rowSums())

######

#Richness by site

#Calculate SR by site
Bird_PA_data_wide_SR_site <- Bird_PA_data_wide_SR %>% 
  group_by(siteid, siteid_star, estuaryname, estuarytype, year, season) %>% 
  summarise(across('Horned Grebe':'Hermit Thrush', sum)) %>% 
  ungroup() %>% 
  #change numeric values to 1 so that it's PA
  mutate(across(c('Horned Grebe':'Hermit Thrush'), ~ if_else(. > 0, 1, 0))) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Horned Grebe':'Hermit Thrush') %>% 
      rowSums())

Bird_SR_site_plot <- Bird_PA_data_wide_SR_site %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  ggplot(aes(x=siteid_star, y=species_richness, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Bird species richness") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) + 
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

Bird_SR_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Birds/Bird_SpeciesRichness_bysite.jpg", 
           Bird_SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Bird_SR_site_metric <- Bird_PA_data_wide_SR_site %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, species_richness) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(species_richness, 4)) %>% 
  #add metric name
  mutate(metric = "bird richness") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(species_richness),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 


```

# Inverts

```{r Data}

#we'll look at the macroinvertebrate data from the large cores and the epifauna data
#this will form the macro-invertebrates

macro_invert_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_benthiclarge_abundance")

epifauna_data <- read.csv("https://nexus.sccwrp.org/empachecker/export?tablename=tbl_epifauna_data")

macro_invert_data_clean <- macro_invert_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, samplelocation, scientificname, live_dead,
         shell_type, unknown_replicate, abundance, region:estuarytype) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))

epifauna_data_clean <- epifauna_data %>% 
  mutate(date = ymd(samplecollectiondate)) %>% 
  #just keep the date, no time
  mutate(date = date(date)) %>% 
  #select important columns 
  select(siteid, estuaryname, date, stationno, transectreplicate, plotreplicate,
         burrows, scientificname, commonname, status, 
         enteredabundance, quadratsize, estimatedabundance, region:estuarytype) %>% 
## add season 
  mutate(
    year = year(date),
    month = month(date),
    season = case_when(
      month %in% c(8, 9, 10, 11, 12) ~ "Fall",
      month %in% c(3, 4, 5, 6) ~ "Spring",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-month) %>% 
  #select only SM BAY data
  filter(year %in% c("2023", "2024")) %>% 
  filter(siteid %in% c("SC-GOL", "SC-VEN", "SC-MAL", "SC-NEW", "SC-BAT",
                       "SC-TOP", "SC-ZUM", "SC-ASE", "SC-BSC"))
  

```

##Large

### Community - large

```{r PA join}

#join the data together for PA
PA_macro_inverts <- macro_invert_data_clean %>%
  #add star so can flag in graphs where sites are open
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  filter(year == 2024) %>% 
  #remove Not recorded
  filter(scientificname != "Not recorded") %>% 
  #filter out unknowns
  filter(!grepl("unknown", scientificname, ignore.case = TRUE)) %>% 
  mutate(presence_absence = ifelse(abundance != 0, 1, 0)) %>% 
  #add column that identifies the data type
  mutate(data_type = "large_inverts")

PA_epifauna <- epifauna_data_clean %>%
  #add star so can flag in graphs where sites are open
  mutate(star = if_else(estuarytype == "Temporarily Closed", "*", "")) %>% 
  mutate(siteid_star = paste0(siteid, star)) %>% 
  filter(year == 2024) %>% 
  #remove Not recorded
  filter(scientificname != "Not recorded") %>% 
  #filter out unknowns
  filter(!grepl("unknown", scientificname, ignore.case = TRUE)) %>% 
  mutate(presence_absence = ifelse(enteredabundance != 0, 1, 0)) %>% 
  #add column that identifies the data type
  mutate(data_type = "epifauna")

#we are going to rbind the two datasets

PA_all <- PA_macro_inverts %>% 
  select(siteid, siteid_star, estuaryname, date, stationno, scientificname,
         presence_absence, year, season, estuarytype, data_type) %>% 
  #we are getting rid of live and dead and location, therefore get rid of dups
  unique() %>% 
  rbind(., PA_epifauna %>% 
          select(siteid, siteid_star, estuaryname, date, stationno, scientificname,
         presence_absence, year, season, estuarytype, data_type) %>% 
          unique())
  
```

```{r heatmap}

## find order of PA from most to least for ordering of heat map
SpeciesOrder <- PA_all %>%
  filter(presence_absence == 1) %>% 
  select(scientificname, siteid) %>% 
  distinct() %>% 
  group_by(scientificname) %>% 
  summarise(Count = n()) %>% 
  arrange(desc(Count))

## get the ordered list of species by their count
SpeciesList <- SpeciesOrder %>% pull(scientificname)

## summarize df and create color category for PA heat map. will add new color if present in both years 
PA_summarized <- PA_all %>%
  group_by(siteid, siteid_star, scientificname) %>%
  summarise(color_category = case_when(
    sum(presence_absence == 1 & season == "Fall") > 0 & 
      sum(presence_absence == 1 & season == "Spring") > 0 ~ "Both",
    sum(presence_absence == 1 & season == "Spring") > 0 ~ "Spring",
    sum(presence_absence == 1 & season == "Fall") > 0 ~ "Fall",
    TRUE ~ "None"
  )) %>%
  ungroup() #

## order based on site order
PA_filter_order <-PA_summarized %>% 
  mutate(siteid_star =  factor(siteid_star, levels = site_order)) %>%
  arrange(siteid_star)  %>% 
  #order based on species count
  mutate(scientificname =  factor(scientificname, levels = SpeciesList)) %>%
  arrange(scientificname)  

## define colors for color category
colors <- c("Spring" = "#DDCC77", "Fall" = "#88CCEE", "Both" = "#117733",
            "None" = "white")

##GGPLOT

n_species <- length(unique(PA_filter_order$scientificname))

heatmap_large_invert <- PA_filter_order %>% 
  ggplot(aes(scientificname, siteid, fill = color_category))+
  geom_tile(colour = "gray50") + # tile border
  scale_fill_manual(values = colors, name = "Season") +
  coord_equal(expand = 0) +
  #scale_y_discrete(limits = rev(levels(fish_filtered_data$siteid_star)), drop = FALSE) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(size = 10, angle = 45, hjust = 1, face = "bold"), # bold common names
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12) 
        
  )+
  labs(x = "Macro-invertebrate species", y = "SiteID", title = "2024 Macro-Invertebrate Presence")+
  guides(fill = guide_legend(override.aes = list(alpha = 1)))+
  #this adds make in the sites with missing data
  geom_hline(yintercept = seq(0.5, length(unique(PA_filter_order$siteid)) - 0.5, by = 1), #add normal grid lines
             color = "black", linewidth = 0.5) +
  geom_vline(xintercept = seq(0.5, n_species - 0.5, by = 1),
             color = "black", linewidth = 0.5)

## print it
heatmap_large_invert

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Macro_Invert_PA_heatmap.jpg", 
           heatmap_large_invert,
       width = 12,
       height = 8,
           dpi = 150)

```

### Metrics - large

```{r species richness}

str(PA_all)
str(PA_all$presence_absence)

Macro_PA_data_wide_SR <- PA_all %>% 
  select(siteid, siteid_star, date, stationno, year, season, estuaryname, 
         estuarytype, scientificname, presence_absence) %>% 
  #get rid of dups bc we don't care where the data comes from anymore
  unique() %>% 
  #Switch to wide format
  pivot_wider(names_from = scientificname, 
              values_from = presence_absence,
              values_fill = 0) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Siliqua lucida':'Ficopomatus enigmaticus') %>% 
      rowSums())

######

#Richness by site

#Calculate SR by site
MacroPA_data_wide_SR_site <- Macro_PA_data_wide_SR %>% 
  group_by(siteid, siteid_star, estuaryname, estuarytype, year, season) %>% 
  summarise(across('Siliqua lucida':'Ficopomatus enigmaticus', sum)) %>% 
  ungroup() %>% 
  #change numeric values to 1 so that it's PA
  mutate(across(c('Siliqua lucida':'Ficopomatus enigmaticus'),
                ~ if_else(. > 0, 1, 0))) %>% 
  #calculate SR
  mutate(species_richness = select(.,
                                  'Siliqua lucida':'Ficopomatus enigmaticus') %>% 
      rowSums())

Macro_SR_site_plot <- MacroPA_data_wide_SR_site %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(species_richness = NA))  %>% 
  ggplot(aes(x=siteid_star, y=species_richness, fill = season)) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Macro-invertebrate species richness") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) + 
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

Macro_SR_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Macroinvert_SpeciesRichness_bysite.jpg", 
           Macro_SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)

## Quartiles

Macro_SR_site_metric <- MacroPA_data_wide_SR_site %>% 
  select(siteid, siteid_star, estuaryname, estuarytype, year, 
         season, species_richness) %>%
  group_by(season) %>% 
  mutate(quartile = ntile(species_richness, 4)) %>% 
  #add metric name
  mutate(metric = "macro-invert richness") %>% 
  group_by(season) %>%
  mutate(
    pct_rank = percent_rank(species_richness),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  ) 

```

## Small

### Figures

```{r richness}

#calling in reduced file from CW for now, need to get data into portal

invert_small_data <- read.csv("/Users/janw/Desktop/projects/Prop50-SMB/Data/CW_invert_counts.csv")

invert_small_data <- invert_small_data %>% 
    #add other site info
  left_join(., Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype) %>% 
              unique()) #%>% 
#   mutate(date = mdy(samplecollectiondate)) %>% 
#   #just keep the date, no time
#   mutate(date = date(date)) %>% 
# ## add season 
#   mutate(
#     year = year(date),
#     month = month(date),
#     season = case_when(
#       month %in% c(8, 9, 10, 11, 12) ~ "Fall",
#       month %in% c(3, 4, 5, 6) ~ "Spring",
#       TRUE ~ NA_character_
#     )
#   ) 

#species richness by core

invert_SR_core_plot <- invert_small_data %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(richness = NA))  %>% 
  ggplot(aes(x=siteid_star, y=richness, 
             fill = season
             )) + 
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), width = 0.6) +
  geom_jitter(aes(color = season, group = season),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 2, alpha = 0.7) +
  ylab("Benthic invert species richness by core") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  scale_color_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

invert_SR_core_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Benthic_SpeciesRichness_bycore.jpg", 
           invert_SR_core_plot,
       width = 10,
       height = 8,
           dpi = 150)

#species richness by site

invert_small_data_site <- read.csv("/Users/janw/Desktop/projects/Prop50-SMB/Data/CW_invert_counts_site.csv")

invert_small_data_site <- invert_small_data_site %>% 
    #add other site info
  left_join(., Site_df %>% 
              select(siteid, siteid_star, estuaryname, estuarytype) %>% 
              unique()) #%>% 
#   mutate(date = mdy(samplecollectiondate)) %>% 
#   #just keep the date, no time
#   mutate(date = date(date)) %>% 
# ## add season 
#   mutate(
#     year = year(date),
#     month = month(date),
#     season = case_when(
#       month %in% c(8, 9, 10, 11, 12) ~ "Fall",
#       month %in% c(3, 4, 5, 6) ~ "Spring",
#       TRUE ~ NA_character_
#     )
#   ) 


invert_SR_site_plot <- invert_small_data_site %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(total_richness = NA))  %>% 
  ggplot(aes(x=siteid_star, y=richness, 
             fill = season
             )) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Benthic invert species richness by site") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

invert_SR_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Benthic_SpeciesRichness_bysite.jpg", 
           invert_SR_site_plot,
       width = 10,
       height = 8,
           dpi = 150)



```

```{r abundance}

#abundance by core

invert_N_core_plot <- invert_small_data %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(abundance = NA))  %>% 
  ggplot(aes(x=siteid_star, y=abundance, 
             fill = season
             )) + 
  geom_boxplot(outlier.shape = NA, 
               position = position_dodge(width = 0.8), width = 0.6) +
  geom_jitter(aes(color = season, group = season),
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              size = 2, alpha = 0.7) +
  ylab("Benthic invert abundance by core") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  scale_color_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

invert_N_core_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Benthic_abundnace_bycore.jpg", 
           invert_N_core_plot,
       width = 10,
       height = 8,
           dpi = 150)

# abundance by site

invert_abundance_site_plot <- invert_small_data_site %>% 
  mutate(season = factor(season, levels = c("Spring", "Fall"))) %>% 
  mutate(siteid_star = factor(siteid_star, levels = site_order_star)) %>% 
  #added completet to fill in missing combinations to reserve bra position
  complete(siteid_star, year, season, fill = list(abundance = NA))  %>% 
  ggplot(aes(x=siteid_star, y=abundance, 
             fill = season
             )) + 
  geom_bar(stat = "identity", color = "black",
           position = position_dodge()) +
  ylab("Benthic invert abundance by site") + 
  xlab("Site") +
  theme_bw() +
  theme(axis.title.x = element_text(face="bold", size=16),
        axis.text.x  = element_text(size=14, angle = 45,
                                    vjust = 1, hjust = 1)) +
  theme(axis.title.y = element_text(angle=90, face="bold", 
                                    size=16, 
                                    vjust=1.2),
        axis.text.y  = element_text(size=14)) +
  scale_fill_manual(name = "Season",
                    values = c("#DDCC77", "#88CCEE"),
                    breaks = c("Spring", "Fall")) +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(colour="black", size = 14)) +
  theme(legend.title = element_text(colour="black", size = 16, face = "bold")) +
  theme(strip.text.x = element_text(size=14, face = "bold")) 

invert_abundance_site_plot

ggsave("/Users/janw/Desktop/projects/Prop50-SMB/Figures/Inverts/Benthic_abundance_bysite.jpg", 
           invert_abundance_site_plot,
       width = 10,
       height = 8,
           dpi = 150)



```


### Metrics

```{r richness}

## Quartiles

Benthic_invert_SR_site_metric <-invert_small_data_site %>% 
  group_by(season) %>% 
  mutate(quartile = ntile(richness, 4)) %>% 
  #add metric name
  mutate(metric = "benthic-invert richness") %>% 
  #group_by(season) %>%
  mutate(
    pct_rank = percent_rank(richness),
    quartile_pct = case_when(
      is.na(pct_rank) ~ NA_integer_,
      pct_rank <= 0.25 ~ 1L,
      pct_rank <= 0.50 ~ 2L,
      pct_rank <= 0.75 ~ 3L,
      TRUE ~ 4L
    )
  )

```

### BPJ scores

```{r BPJ score summary}

setwd("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/data-inventory")
getwd()
bpj.scores <- read.csv("BPJ_scores.csv")
head(bpj.scores)

# Pivot data
bpj.scores.long <- bpj.scores %>%
  pivot_longer(cols = c(Avg.Rnd.1, Avg.Rnd.2),
               names_to = "Round",
               values_to = "Average") %>%
  mutate(stdev = ifelse(Round == "Avg.Rnd.1", stdev_R1, stdev.2_R2))

round.colors = c("Avg.Rnd.1" = "darkblue", "Avg.Rnd.2" = "#00cCCC")
class.colors = c("seasonally dry, small lagoons" = "#330000", "sandy beach, usually wet lagoons" = "#FFCCCC" , "larger lagoons, usually wet" = "#990000")

# Plot 1: plot just stdev in both rounds

scoring_summary1 <- ggplot(bpj.scores.long, aes(x = Estuary, y = Average, color = Round))  + 
  geom_tile(aes(y = 0.5, fill = Class),
            height = 0.3, color = NA) +
  geom_point(size =  3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = Average - stdev,
                    ymax = Average + stdev),
                width = 0.1, position = position_dodge(width = 0.4))  +
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.05)), breaks = 1:5) +
  scale_color_manual(values = round.colors, labels = c("Avg.Rnd.1" = "Round 1", "Avg.Rnd.2" = "Round 2"), name = "Panel Review") +
  scale_fill_manual(values = class.colors, name = "Estuary class") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 14),
        legend.position = "right") + 
  labs(y = "Score 1 -5", x = "Estuary") 
  
scoring_summary1

ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/SOTB_ppt/scoring_summary1.jpg", 
           scoring_summary1,
       width = 8,
       height = 8,
           dpi = 300)


# Plot 2: Plot side-by side barplot with consensus 
#repivot data to include consensus
bpj.scores.consensus <- bpj.scores %>%
  pivot_longer(cols = c(Avg.Rnd.1, Avg.Rnd.2, Consensus),
               names_to = "Round",
               values_to = "Average") %>%
  mutate(stdev = case_when(
    Round == "Avg.Rnd.1" ~ stdev_R1,
    Round == "Avg.Rnd.2" ~ stdev.2_R2,
    Round == "Consensus" ~ NA_real_
  ))

round.colors = c("Avg.Rnd.1" = "darkblue", "Avg.Rnd.2" = "#00cCCC", "Consensus" = "#339933")
class.colors = c("seasonally dry, small lagoons" = "black", "sandy beach, usually wet lagoons" = "grey90" , "larger lagoons, usually wet" = "grey50")


consensus_summary1 <-  ggplot(bpj.scores.consensus, aes(x = Estuary, y = Average, color = Round))  + 
  geom_tile(aes(y = 0.5, fill = Class),
            height = 0.3, color = NA) +
  geom_point(size =  3, position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = Average - stdev,
                    ymax = Average + stdev),
                width = 0.1, position = position_dodge(width = 0.4))  +
  scale_y_continuous(expand = expansion(mult = c(0.2, 0.05)), breaks = 1:5) +
  scale_color_manual(values = round.colors, labels = c("Avg.Rnd.1" = "Round 1", "Avg.Rnd.2" = "Round 2", "Consensus" = "Consensus"), name = "Panel Review") +
  scale_fill_manual(values = class.colors, name = "Estuary class") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 14),
        legend.position = "right") + 
  labs(y = "Score 1 -5", x = "Estuary") 

  
  consensus_summary1
  
  ggsave("/Users/jilltupitza/Library/CloudStorage/OneDrive-SCCWRP/SM-Bay/SOTB_ppt/scoring_summary2.jpg", 
           consensus_summary1,
       width = 8,
       height = 8,
           dpi = 300)

# Slide 23 replacement

```

